Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2025-03-11 report/emse24-response.tex
chore: update emse24 revised documents

  By analyzing whether the findings hold consistently across these clusters, the authors could find patterns specific to notebook complexity. This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity. 
  % TODO: we explored this idea but we observed that large notebooks could be (uninteresting) eg. notebook X while smaller... link this with SS (gives similar weightable to all FMs without making assumptions about complexity or other properties)

  Including this stratified analysis would significantly strengthen the study's contributions.
\end{revcomment}

% TODO also add the benefits of using stratified sampling to ensure diversity in population of FMs; "while not perfect, we use SS to increase diversity of our candidates"
% TODO add example of a really large notebook with simple FM and a small one with complex FM
\Us Thank you for your feedback. We have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work. While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity. We agree that the number of cells (or size) of a notebook is a valid albeit not the only metric that should be considered to define notebook complexity.

@@

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. We renamed the title of section 3.2 Case Studies to \emph{Exploratory Case Study Analysis} to be verbose about the methodology. We added more details regarding the \emph{site} and the \emph{selection criteria} of the study. Furthermore, details regarding the sampling technique, the bias introduced, and our mitigation strategy has also been added. Finally, to validate the taxonomy of feedback mechanisms, the second author independently annotated all candidates using the codes that were generated during the first round of annotations. We have reported the inter-rater agreement which was calculated using the \emph{Cohen's Kappa} metric in the manuscript.

@@

  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). 
  % TODO: after introducing TF-IDF add why this is useful for us; it allows us to find not only common FMs but also specialized ones that address corner cases
  This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

@@

% TODO: our goal was not to create an extensive taxonomy of FMs, but to validate if there is something worth future research; this supports the use of a time budget for the analysis
% TODO: in order to scale analysis to a generalizable population, we need automated tools; our methodology does not scale; smaller but insightful analysis; to scale we need to trade-off between deep insights and generalizability
We have also revised section 6 Threats to Validity (Internal Validity) to discuss the pros and cons of using a fixed time resource. While we agree that the use of a fixed time resource has limitations, other stopping criteria, such as those based on theoretical saturation, are also subject to bias from the annotator. In contrast, a fixed time resource reflects the practical resource constraints that researchers face in the real world.

>>>>

  By analyzing whether the findings hold consistently across these clusters, the authors could find patterns specific to notebook complexity. This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity. Including this stratified analysis would significantly strengthen the study's contributions.
\end{revcomment}

\Us Thank you for your feedback. We explored the idea of using notebook size (number of cells) as a proxy for complexity. However, we found that larger notebooks don't necessarily contain more complex feedback mechanisms, and smaller notebooks can sometimes contain sophisticated feedback constructs. For example, the largest notebook in our dataset contains over 500 simple print statements (see large-notebook.ipynb: \url{https://surfdrive.surf.nl/files/index.php/s/FDQkCFxQNAVZIcq}), while some compact notebooks implemented complex assertion patterns with custom validation logic (see small-notebook.ipynb: \url{https://surfdrive.surf.nl/files/index.php/s/FDQkCFxQNAVZIcq}).

To avoid making assumptions about the relationship between notebook size and feedback mechanism complexity, we opted for stratified random sampling based on TF-IDF values. This approach ensures a diverse population of feedback mechanisms in our analysis, giving appropriate representation to both common patterns and specialized implementations regardless of the notebook size.

Additionally, we have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work. While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity. We agree that the number of cells (or size) of a notebook is a valid albeit not the only metric that should be considered to define notebook complexity.

@@

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. We renamed the title of section 3.2 Case Studies to \emph{Exploratory Case Study Analysis} to be verbose about the methodology. We added more details regarding the \emph{site} and the \emph{selection criteria} of the study. Furthermore, details regarding the sampling technique using TF-IDF, the bias introduced, and our mitigation strategy using stratified random sampling has also been added.

Finally, to validate the taxonomy of feedback mechanisms, the second author independently annotated all candidates using the codes that were generated during the first round of annotations. We have reported the inter-rater agreement which was calculated using the \emph{Cohen's Kappa} metric in the manuscript.

@@

  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This technique allows us to surface both frequently occurring feedback mechanisms and those that are rarer in the dataset.

  This normalization using IDF assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

@@

We have also revised section 6 Threats to Validity (Internal Validity) to discuss the pros and cons of using a fixed time resource. While we agree that the use of a fixed time resource has limitations, other stopping criteria, such as those based on theoretical saturation, are also subject to bias from the annotator. In contrast to other methods, a fixed time resource reflects the practical resource constraints that researchers face in the real world. Furthermore, our goal for the study was not to create an extensive taxonomy of feedback mechanisms but to explore if this direction of research warrants future work, and the use of a time budget for the analysis also supports this goal.

### 2025-03-11 report/emse24-revised.tex
chore: update emse24 revised documents

The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \lstinline{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

>>>>

The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This technique allows us to surface both frequently occurring feedback mechanisms and those that are rarer in the dataset.

The normalization using IDF assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \lstinline{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.
